Investigating Neolithic pottery usage in the Yanyuan Basin in Southwest China using organic residue analysis

### 2023-01-14 analysis/paper/paper.Rmd
added details of GC-FID to the method section

<!-- sites and research questions-->
As an early Neolithic site, Guijiabao (ca. 4800-4300 BP) is crucial to understanding cultural interactions and the spread of crops into the Yanyuan Basin due to its location at the intersection of the middle Hengduan Mountain ranges, the Sichuan Plateau, and the Yunnan-Guizhou Plateau. The recent evidence at Guijiabao reveals millet as the main crop with little rice according to the results of flotation [@citation], different from the abundant rice grains found at other Neolithic sites in that area. Although the plant remains suggest the possible domestication of millet in the early Neolithic, we still lack direct evidence of culinary practices during the Neolithic transition. To explore the extent of acceptance of new crops and species in the Yanyuan Basin, we ask the questions: What was the culinary practices during the Neolithic transition? Could we identify the evidence of new crops or species incorporated into the local dietary systems in the Early Neolithic period? We answer those questions by examining potsherds collected from Guijiabao using organic residue analysis, which could offer a better understanding of food sources during the Neolithic transition in the Yanyuan Basin.

@@

# Geographical and Cultural Background of Guijiabao in Yanyuan Basin
<!--Info about Yanyuan area-->
The Yanyuan area in Sichuan province is a pull-apart basin located in the middle eastern part of the Hengduan Mountain, which is a group of parallel mountain ranges in a north-south direction from the Tibetan Plateau to Yunnan-Guizhou Plateau. Although the ranges isolate areas with their own unique geographic conditions and cultures, river valleys provide pathways for the movements of people and the exchange of goods between cultural groups since the prehistoric period [@Lin2022; @Hao2022]. The archaeological sites in the Yanyuan Basin, therefore, are important to understanding the cultural interactions between southwest China and other regions, including northwest China, the Sichuan area, and northwest Yunnan.

@@

The remaining 95% aliquot was fractioned by solid phase extraction (Aminopropyl, NH2, column with 0.5 g sorbent) with 8 mL dichloromethane:Isopropyl alcohol (3:1), 8 mL 4% acetic acid in ethyl ether, and 6 mL methanol respectively. The second fraction was methylated to produce fatty acid methyl esters (FAMEs) using 1 mL dry hexane and 2 mL 10% (v/v) acetyl chloride in dry methanol, and heated at 70 degree C overnight. The FAME derivatives were extracted with hexane ($3\times10$ mL) and blown down with nitrogen stream. About 5% aliquot dissolved in toluene was analyzed by GC-FID for quantification and the remaining was analyzed by GC-C-IRMS.

The GC-MS analysis was performed using an Agilent 7683 B series gas chromatograph attached to an Agilent 5975 inert mass selective detector. Samples were injected in splitless mode at 320 °C and compounds were separated with a VF-17ms column (60 m length, 0.25 mm i.d., 0.25 μm film thickness). Helium carrier is in constant flow rate of 1.3 ml/min. The temperature of the oven was started from 90 °C for 1 minute and increased at a rate of 5 °C/min to 320 °C, where it was held for 23 min. The GC-FID analysis for the FAME samples was carried out using Agilent 6890N gas chromatograph with an FID detector. The GC-C-IRMS analyses of FAMEs of individual fatty acids were conducted using a Thermo Delta V Plus mass spectrometer linked to a GC Trace 1310 through an Isolink Interface. The sample injection volume was 1 ul in toluene and was injected in splitless mode. Separation was carried out using the same model of column for GC-MS and the GC oven temperature was set from 90 °C for 1.5 minutes, and then raised at a rate of 5 °C/min to 320 °C, with an 18-minute hold. Helium was used as a gas carrier in constant flow mode at 1.24 ml/minute.

@@

```{r FA-composition-table, cache= TRUE}
quant_FA <-
  readxl::read_excel(here::here("analysis", "data", "raw_data", "210824_FAMES_pot_LW_publication.xlsx"), 
                     skip = 2, col_names = TRUE)

@@

1000 BC horse
Gaushan-90 pig, plant-most are rice: 95% and a few millet

>>>>

<!-- sites and research questions-->
As an early Neolithic site, Guijiabao (ca. 4800-4300 BP) is crucial to understanding cultural interactions and the spread of crops into the Yanyuan Basin due to its location at the intersection of the middle Hengduan Mountain ranges, the Sichuan Basin, and the Yunnan-Guizhou Plateau. The recent evidence at Guijiabao reveals millet as the main crop with little rice according to the results of flotation [@citation], different from the abundant rice grains found at other Neolithic sites in that area. Although the plant remains suggest the possible domestication of millet in the early Neolithic, we still lack direct evidence of culinary practices during the Neolithic transition. To explore the extent of acceptance of new crops and species in the Yanyuan Basin, we ask the questions: What was the culinary practices during the Neolithic transition? Could we identify the evidence of new crops or species incorporated into the local dietary systems in the Early Neolithic period? We answer those questions by examining potsherds collected from Guijiabao using organic residue analysis, which could offer a better understanding of food sources during the Neolithic transition in the Yanyuan Basin.

@@

# Geographical and Cultural Background of Guijiabao in Yanyuan Basin
<!--Info about Yanyuan area-->
The Yanyuan area in Sichuan province is a pull-apart basin located in the middle eastern part of the Hengduan Mountain, which is a group of parallel mountain ranges in a north-south direction from the Tibetan Plateau to Yunnan-Guizhou Plateau. Although the ranges isolate areas with their own unique geographic conditions and cultures, river valleys provide pathways for the movements of people and the exchange of goods between cultural groups since the prehistoric period [@Lin2022; @Hao2022]. Therefore, the archaeological sites in the Yanyuan Basin are important to understanding the cultural interactions between southwest China and other regions, including northwest China, the Sichuan area, and northwest Yunnan.

@@

The remaining 95% aliquot was fractioned by solid phase extraction (Aminopropyl, NH2, column with 0.5 g sorbent) with 8 mL dichloromethane:Isopropyl alcohol (3:1), 8 mL 4% acetic acid in ethyl ether, and 6 mL methanol respectively. The second fraction was methylated to produce fatty acid methyl esters (FAMEs) with 1 mL dry hexane and 2 mL 10% (v/v) acetyl chloride in dry methanol, and heated at 70 degree C overnight. The FAME derivatives were extracted with hexane ($3\times10$ mL) and blown down with nitrogen stream. About 5% aliquot dissolved in toluene was analyzed by GC-FID for quantification and the remaining was analyzed by GC-C-IRMS.

The GC-MS analysis was performed using an Agilent 7683 B series gas chromatograph attached to an Agilent 5975 inert mass selective detector. Samples were injected in splitless mode at 320 °C and compounds were separated with a VF-17ms column (60 m length, 0.25 mm i.d., 0.25 μm film thickness). Helium carrier is in constant flow rate of 1.3 ml/min. The temperature of the oven was started from 90 °C for 1 minute and increased at a rate of 5 °C/min to 320 °C, where it was held for 23 min. The GC-FID analysis for the FAME samples was carried out using Agilent 6890N gas chromatograph equipped with a PTV inlet in splitless mode. The carrier gas was helium at 1.6 ml/min. The GC system was an Agilent 6890N equipped with a programmed temperature vaporization (PTV) inlet in splitless mode. The GC oven was held at 90°C for 1.3 minutes, and then raised to 320°C at a rate of 5°C/min. The final temperature was held for 22 minutes. The GC-C-IRMS analyses of FAMEs of individual fatty acids were conducted using a Thermo Delta V Plus mass spectrometer linked to a GC Trace 1310 through an Isolink Interface. The sample injection volume was 1 ul in toluene and was injected in splitless mode. Separation was carried out using the same model of column for GC-MS and the GC oven temperature was set from 90 °C for 1.5 minutes, and then raised at a rate of 5 °C/min to 320 °C, with an 18-minute hold. Helium was used as a gas carrier in constant flow mode at 1.24 ml/minute.

@@

```{r FA-composition-table, cache= TRUE, cache.extra=tools::md5sum(here::here("analysis/data/raw_data/210824_FAMES_pot_LW_publication.xlsx"))}
quant_FA <-
  readxl::read_excel(here::here("analysis", "data", "raw_data", "210824_FAMES_pot_LW_publication.xlsx"), 
                     skip = 2, col_names = TRUE)

@@

net sinkers
1000 BC horse
Gaushan-90 pig, plant-most are rice: 95% and a few millet
